\section{Threat Hunting Fundamentals}

Threat hunting is the proactive search for adversary activity that has evaded automated detection. 
Unlike alert triage, hunting begins without a predefined signal; instead, the analyst forms hypotheses 
based on telemetry, threat intelligence, and knowledge of system behavior. The goal is to identify subtle 
patterns, weak signals, and behavioral anomalies that indicate malicious activity. This chapter introduces 
the core mindset and techniques used in modern threat hunting, supported by SPL and KQL examples that reflect 
real investigative workflows.

% --------------------------------------------------------------------
% SECTION 1: The Threat Hunting Mindset
% --------------------------------------------------------------------

\subsection*{The Threat Hunting Mindset}

Threat hunting is driven by curiosity, pattern recognition, and an understanding of how adversaries operate. 
A hunt typically begins with a hypothesis. This hypothesis can be driven by:

\begin{itemize}
    \item \textbf{Threat intelligence} — “A threat actor is known to use encoded PowerShell commands.”
    \item \textbf{Telemetry anomalies} — “There is a spike in failed logons.”
    \item \textbf{Environmental knowledge} — “This server should never initiate outbound connections.”
\end{itemize}

Once a hypothesis is defined, the hunter uses SPL or KQL to explore relevant telemetry, pivot across related 
events, and refine the search until meaningful patterns emerge. This iterative process is what separates hunting 
from simple querying.

% --------------------------------------------------------------------
% SECTION 2: Core Hunting Techniques
% --------------------------------------------------------------------

\subsection*{Core Hunting Techniques}

Threat hunting relies on several foundational techniques:

\begin{itemize}
    \item \textbf{Filtering} — Narrowing down large datasets to relevant activity.
    \item \textbf{Aggregation} — Grouping events to identify unusual patterns.
    \item \textbf{Rarity analysis} — Finding activity that deviates from normal behavior.
    \item \textbf{Behavioral correlation} — Connecting related events across multiple data sources.
\end{itemize}

These techniques form the basis of most hunting workflows.

% --------------------------------------------------------------------
% SECTION 3: Process Creation Hunting
% --------------------------------------------------------------------

\subsection*{Process Creation Hunting}

One of the most common starting points in a hunt is process creation activity. Sysmon Event ID 1 provides detailed 
insight into how processes are launched, including parent-child relationships and command-line arguments. The 
following example searches for suspicious PowerShell usage, a frequent indicator of malicious activity.

\textbf{SPL Example}

\begin{lstlisting}
index=sysmon EventCode=1 Image="*powershell.exe"
| search CommandLine="*EncodedCommand*"
| table _time, Computer, User, ParentImage, Image, CommandLine
\end{lstlisting}

\textbf{KQL Equivalent}

\begin{lstlisting}
Sysmon
| where EventID == 1 and Image endswith "powershell.exe"
| where CommandLine contains "EncodedCommand"
| project TimeGenerated, Computer, User, ParentImage, Image, CommandLine
\end{lstlisting}

% --------------------------------------------------------------------
% SECTION 4: Rarity Analysis
% --------------------------------------------------------------------

\subsection*{Rarity Analysis}

Adversaries often stand out not because their activity is obviously malicious, but because it is uncommon compared 
to normal system patterns. The following example highlights rare parent-child process pairs.

\textbf{SPL Example}

\begin{lstlisting}
index=sysmon EventCode=1
| stats count by ParentImage, Image
| where count < 3
\end{lstlisting}

\textbf{KQL Equivalent}

\begin{lstlisting}
Sysmon
| where EventID == 1
| summarize Count=count() by ParentImage, Image
| where Count < 3
\end{lstlisting}

% --------------------------------------------------------------------
% SECTION 5: Authentication Hunting
% --------------------------------------------------------------------

\subsection*{Authentication Hunting}

Authentication activity is another rich source of hunting leads. Repeated failed logons, unusual logon types, or 
logons from unexpected hosts can indicate credential misuse or lateral movement. The following example identifies 
accounts with excessive failed logons.

\textbf{SPL Example}

\begin{lstlisting}
index=wineventlog EventCode=4625
| stats count by AccountName, WorkstationName
| where count > 10
\end{lstlisting}

\textbf{KQL Equivalent}

\begin{lstlisting}
SecurityEvent
| where EventID == 4625
| summarize Count=count() by Account, Workstation
| where Count > 10
\end{lstlisting}

% --------------------------------------------------------------------
% SECTION 6: Network Hunting
% --------------------------------------------------------------------

\subsection*{Network Hunting}

Network telemetry plays a critical role in hunting. DNS logs, proxy logs, and NetFlow data can reveal command-and-control 
activity, data exfiltration, or lateral movement. The following example identifies rare outbound connections.

\textbf{SPL Example}

\begin{lstlisting}
index=sysmon EventCode=3
| stats count by DestinationIp, DestinationPort
| where count < 5
\end{lstlisting}

\textbf{KQL Equivalent}

\begin{lstlisting}
Sysmon
| where EventID == 3
| summarize Count=count() by DestinationIp, DestinationPort
| where Count < 5
\end{lstlisting}

% --------------------------------------------------------------------
% SECTION 7: Why These Fundamentals Matter
% --------------------------------------------------------------------

\subsection*{Why These Fundamentals Matter}

Threat hunting is ultimately about pattern recognition and curiosity. The examples in this chapter demonstrate the 
foundational techniques—filtering, aggregation, rarity analysis, and behavioral correlation—that underpin more advanced 
hunts. Later chapters expand these fundamentals into full hunting playbooks and detection patterns that reflect real-world 
adversary behavior.
